---
title: "Heart Disease Dataset – Exploratory Analysis"
author: "Atilul Islam Sajib, Zafrin Sultana, Sakib Hasan, & Mohammad Mahmudul Hasan"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(dplyr)
library(corrplot)

set.seed(123)

check_missing <- function(df) colSums(is.na(df))
class_balance <- function(df) table(df$HeartDisease)
```

# Introduction

This report provides an exploratory data analysis (EDA) of the processed **Heart Disease Training dataset**.

---

# Load Dataset

```{r}
train <- read.csv("../data/processed/train_scaled.csv", stringsAsFactors = TRUE)

print_basic_info <- function(df, target_col) {
  cat("\nData overview\n")
  print(dim(df))
  str(df)
  
  cat("\n\nSummary:\n")
  print(summary(df))
  
  cat("\nMissing values:\n")
  print(check_missing(df))
  
  cat("\nTarget class balance:\n")
  print(class_balance(df))
}

print_basic_info(train, "HeartDisease")

numeric_cols <- c("Age", "RestingBP", "Cholesterol",
                  "FastingBS", "MaxHR", "Oldpeak")

cat_cols <- c("Sex", "ChestPainType", "RestingECG",
              "ExerciseAngina", "ST_Slope")
```

# 1. Hypothesis Testing: Numerical + Categorical features

```{r hypothesis-testing}
##############################################################
#                           T-TEST                         
##############################################################

t_test <- function(feature_name, group_no, group_yes) {

  n_no  <- length(group_no)
  n_yes <- length(group_yes)
  
  mean_no  <- mean(group_no)
  mean_yes <- mean(group_yes)
  
  variance_no  <- var(group_no)
  variance_yes <- var(group_yes)
  
  # t-value
  t_value <- (mean_no - mean_yes) / sqrt(variance_no/n_no + variance_yes/n_yes)
  
  # formula -> from Stat Course
  df <- (variance_no/n_no + variance_yes/n_yes)**2 /
    ((variance_no**2)/(n_no**2 * (n_no - 1)) +
       (variance_yes**2)/(n_yes**2 * (n_yes - 1)))
  
  # p-value
  p_value <- 2 * (1 - pt(abs(t_value), df))
  
  # print all the conclusion
  cat("\n----------------------------------------------------------------\n")
  cat("Feature:", feature_name, "\n\n")
  
  # Hypotheses
  cat("H0(Null Hypothesis): ")
  cat("People with and without heart disease have similar", feature_name, ".\n\n")
  
  cat("Ha(Alternative Hypothesis): ")
  cat("The two groups do NOT have similar", feature_name, ".\n\n")
  
  # Test values
  cat("Test Results:\n")
  cat("  t-value:", round(t_value, 4), "\n")
  cat("  p-value:", format(p_value, scientific = TRUE), "\n\n")
  
  cat("Group Averages:\n")
  cat("  No heart disease:", round(mean_no, 4), "\n")
  cat("  Yes heart disease:", round(mean_yes, 4), "\n\n")
  
  # Decision
  if (p_value < 0.05) {
    cat("Decision:")
    cat("  The test is significant, so We reject the null hypothesis.\n\n")
    
    if (mean_yes > mean_no) {
      cat("Conclusion:")
      cat("  People with heart disease tend to have higher", feature_name, ".\n")
      cat("  This feature seems related to heart disease.\n")
    } else {
      cat("Conclusion:")
      cat("  People without heart disease tend to have higher", feature_name, ".\n")
      cat("  The feature still differs between groups.\n")
    }
    
  } else {
    cat("Decision:")
    cat("  The test is not significant, so We do not reject the null hypothesis.\n\n")
    
    cat("Conclusion:")
    cat("  This feature does not show a meaningful difference between the groups.\n")
  }
}

##############################################################
#                        CHI-SQUARE TEST                     
##############################################################

chi_square <- function(feature_name, feature_vector, target_vector) {
  
  # Create contingency table - Stat Course
  tbl <- table(feature_vector, target_vector)
  observed <- as.matrix(tbl)
  
  # Row totals, column totals, grand total
  row_totals <- rowSums(observed)
  col_totals <- colSums(observed)
  grand_total <- sum(observed)
  
  # Expected counts (row total * col total / grand total)
  expected <- outer(row_totals, col_totals) / grand_total
  
  # Chi-square statistic
  chi_square_value <- sum((observed - expected)**2 / expected)
  
  # Degrees of freedom
  df <- (nrow(observed) - 1) * (ncol(observed) - 1)
  
  # p-value
  p_value <- 1 - pchisq(chi_square_value, df)
  
  # Print all conclusions
  cat("\n----------------------------------------------------------------\n")
  cat("Feature:", feature_name, "\n\n")
  
  # Hypotheses
  cat("H0(Null Hypothesis): ")
  cat("People with and without heart disease have similar", feature_name, "distribution.\n\n")
  
  cat("Ha(Alternative Hypothesis): ")
  cat("The two groups do NOT have similar", feature_name, "distribution.\n\n")
  
  # Test values
  cat("Test Results:\n")
  cat("  Chi-square value:", round(chi_square_value, 4), "\n")
  cat("  p-value:", format(p_value, scientific = TRUE), "\n")
  cat("  Degrees of freedom:", df, "\n\n")
  
  # Decision
  if (p_value < 0.05) {
    cat("Decision:")
    cat("  The test is significant, so we reject the null hypothesis.\n\n")
    
    cat("Conclusion:")
    cat("  This feature seems related to heart disease.\n")
    
  } else {
    cat("Decision:")
    cat("  The test is not significant, so we do not reject the null hypothesis.\n\n")
    
    cat("Conclusion:")
    cat("  This feature does not show a meaningful difference between the groups.\n")
  }
}

############################################################
#              Run the T-Tests for Numeric Features         
############################################################

train <- read.csv("../data/processed/train_scaled.csv", stringsAsFactors = TRUE)

numeric_cols <- c("Age", "RestingBP", "Cholesterol",
                  "FastingBS", "MaxHR", "Oldpeak")

for (col in numeric_cols) {
  group_no  <- train[[col]][train$HeartDisease == "No"]
  group_yes <- train[[col]][train$HeartDisease == "Yes"]
  t_test(col, group_no, group_yes)
}

cat("\n******* Chi-Square Tests for Categorical Features *******\n")

############################################################
#        Run Chi-Square Tests for Each Categorical Feature  
############################################################

train <- read.csv("../data/processed/train_scaled.csv", stringsAsFactors = TRUE)

categorical_cols <- c("Sex", "ChestPainType", "RestingECG",
                      "ExerciseAngina", "ST_Slope")

for (col in categorical_cols) {
  chi_square(col, train[[col]], train$HeartDisease)
}
```

# 2. Distribution of Target class

```{r pie_chart}
pie_data <- train %>% 
  count(HeartDisease) %>% 
  mutate(prop = n / sum(n),
         lbl = paste0(HeartDisease, " (", scales::percent(prop), ")"))

ggplot(pie_data, aes(x = "", y = prop, fill = HeartDisease)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  theme_void() +
  labs(title = "HeartDisease Class Distribution (Pie Chart)") +
  geom_text(aes(label = lbl), position = position_stack(vjust = 0.5))
```

# 3. Distribution of Numeric Features

```{r}
for (col in numeric_cols) {
  print(
    ggplot(train, aes_string(x = col)) +
      geom_histogram(bins = 30, fill = "skyblue", color = "black") +
      theme_minimal() +
      labs(title = paste("Distribution of", col))
  )
}
```

# 4. Distribution of Categorical Features

```{r}
for (col in cat_cols) {
  print(
    ggplot(train, aes_string(x = col)) +
      geom_bar(fill = "orange", color = "black") +
      theme_minimal() +
      labs(title = paste("Distribution of", col))
  )
}
```

# 5. Numeric Features vs Heart Disease

```{r}
for (col in numeric_cols) {
  print(
    ggplot(train, aes_string(x = "HeartDisease", y = col, fill = "HeartDisease")) +
      geom_boxplot() +
      theme_minimal() +
      labs(title = paste(col, "by HeartDisease"))
  )
}
```

# 6. Categorical Features vs Heart Disease

```{r}
for (col in cat_cols) {
  print(
    ggplot(train, aes_string(x = col, fill = "HeartDisease")) +
      geom_bar(position = "fill") +
      theme_minimal() +
      labs(title = paste(col, "vs HeartDisease (Proportion)"),
           y = "Proportion")
  )
}
```

# 7. Correlation Matrix

```{r}
numeric_data <- train[, numeric_cols]
cor_mat <- cor(numeric_data)

corrplot(cor_mat,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         number.cex = 0.7,
         tl.cex = 0.8,
         tl.col = "black")
```

# 8. Scatter Plots (Age vs Key Numeric Features)

```{r}
key_pairs <- c("RestingBP", "Cholesterol", "MaxHR", "Oldpeak")

for (col in key_pairs) {
  print(
    ggplot(train, aes_string(x = "Age", y = col, color = "HeartDisease")) +
      geom_point(alpha = 0.6, size = 2) +
      theme_minimal() +
      labs(
        title = paste("Scatter Plot:", "Age vs", col),
        x = "Age",
        y = col
      )
  )
}
```

# 9. Scatter Plot With Smooth Trend Line

```{r}
ggplot(train, aes(Age, MaxHR, color = HeartDisease)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE) +
  theme_minimal() +
  labs(title = "Age vs MaxHR with Trend Line")
```

# 10. Faceted Scatter Plot using (MaxHR vs Oldpeak by Sex)

```{r}
ggplot(train, aes(MaxHR, Oldpeak, color = HeartDisease)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ Sex) +
  theme_minimal() +
  labs(title = "MaxHR vs Oldpeak (Faceted by Sex)")
```

# 11. Pairplot for Numeric Features

```{r}
ggpairs(
  train[, c(numeric_cols, "HeartDisease")],
  aes(color = HeartDisease, alpha = 0.5)
)
```
